---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.path  = "man/figures/README-",
  out.width = "100%"
)
```

# ggspec

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/clement-lee/ggspec/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/clement-lee/ggspec/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/clement-lee/ggspec/graph/badge.svg)](https://app.codecov.io/gh/clement-lee/ggspec)
[![CRAN status](https://www.r-pkg.org/badges/version/ggspec)](https://CRAN.R-project.org/package=ggspec)
<!-- badges: end -->

`ggspec` extracts the full declarative specification of a `ggplot2` object —
layers, aesthetic mappings, scales, facets, coordinate system, and labels — as
tidy data frames. A second tier of functions enables structural comparison of
two ggplot objects, supporting automated plot testing, auditing, and
framework-agnostic grading workflows.

## Installation

Install the development version from GitHub:

```r
# install.packages("remotes")
remotes::install_github("clement-lee/ggspec")
```

## Usage

### Extracting a spec

```{r usage-spec}
library(ggspec)
library(ggplot2)

p <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~drv) +
  labs(title = "Engine displacement vs highway MPG")

spec_layers(p)
spec_aes(p)
```

### Comparing two plots

```{r usage-equiv}
ref <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  facet_wrap(~drv)

obs_correct <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  facet_wrap(~drv)

obs_wrong <- ggplot(mpg, aes(displ, hwy)) +
  geom_smooth() +
  facet_wrap(~cyl)

equiv_plot(ref, obs_correct)
equiv_plot(ref, obs_wrong)
```

### Enriching a spec with build-derived defaults

`enrich_spec()` uses `ggplot_build()` to identify which parameters and
aesthetics were explicitly set by the user versus filled in by ggplot2:

```{r usage-enrich}
es <- enrich_spec(p)

# Non-aesthetic parameters with explicit flag
es$params_tbl[[1]]

# Aesthetics resolved by ggplot2, with explicit flag
es$built_aes[[1]]
```

## Key functions

| Tier | Function | What it returns |
|---|---|---|
| Extraction | `spec_layers()` | One row per layer |
| Extraction | `spec_aes()` | One row per layer × aesthetic |
| Extraction | `spec_scales()` | One row per scale |
| Extraction | `spec_facets()` | Facet type and variables |
| Extraction | `spec_labels()` | One row per label |
| Extraction | `spec_coord()` | Coordinate system |
| Extraction | `enrich_spec()` | spec_layers + default/explicit flags |
| Comparison | `equiv_plot()` | All checks in one call |
| Comparison | `equiv_layers()` | Geom and stat per layer |
| Comparison | `equiv_aes()` | Aesthetic mappings |
| Comparison | `compare_plots()` | Canonicalise then compare |
| Check | `check_plot()` | Framework-agnostic assertion |
| Check | `expect_equiv_plot()` | testthat expectation |

## Related packages

- **[ggcheck](https://github.com/rstudio/ggcheck)** — designed for
  `learnr`/`gradethis` pipelines; returns ad-hoc objects.
  `ggspec` returns rectangular, pipeable tibbles and has no grading framework
  dependency.

## License

MIT
