---
title: "Equivalence Patterns and Canonicalisation Modes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Equivalence Patterns and Canonicalisation Modes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Introduction

Two ggplot2 calls can produce identical visual output while being written in
structurally different ways — different placement of `aes()`, different geoms
with built-in stat computation vs pre-computed data, or `coord_flip()` instead
of swapped aesthetics. This vignette catalogues the equivalence patterns
supported by ggspec, documents which function or mode is needed for each, and
flags patterns not yet covered.

The patterns fall into three tiers:

| Tier | Entry point | Key rules applied |
|---|---|---|
| Direct | `equiv_plot()` | Resolved aes, geom-name subset |
| Structural | `compare_plots(mode = "structural")` | `geom_col` → `geom_bar`, layer order |
| Visual | `compare_plots(mode = "visual")` | `coord_flip` absorbed, scale `name` → `labs()` |

```{r}
library(ggspec)
library(ggplot2)
library(dplyr)
```

---

## 1. Inheritance transparency

ggspec resolves aesthetic mappings (via `spec_aes(inherit = "resolve")`) before
any comparison, so different coding styles that yield the same final mapping all
pass `equiv_plot()` directly.

### 1.1 Global, local, and mixed `aes()`

```{r}
# Global aes — canonical form
p_global <- ggplot(airquality, aes(x = Day, y = Wind)) +
  geom_line() + geom_point()

# All aesthetics local to each layer
p_local <- ggplot(airquality) +
  geom_line(aes(x = Day, y = Wind)) +
  geom_point(aes(x = Day, y = Wind))

# Mixed: global x, local y per layer
p_mixed <- ggplot(airquality, aes(x = Day)) +
  geom_line(aes(y = Wind)) +
  geom_point(aes(y = Wind))
```

```{r}
equiv_plot(p_global, p_local, check = c("layers", "aes"))
equiv_plot(p_global, p_mixed, check = c("layers", "aes"))
```

All three pass. The `source` column in `spec_aes()` records where each mapping
originates (`"global"`, `"local"`, or `"resolved"`), but the comparison only
examines the resolved variable name.

### 1.2 Global vs per-layer data

Specifying data globally in `ggplot()` vs attaching it to individual layers is
also transparent:

```{r}
# Global data
p_data_global <- ggplot(airquality, aes(x = Day, y = Wind)) +
  geom_line() + geom_point()

# Per-layer data
p_data_local <- ggplot() +
  geom_line(aes(x = Day, y = Wind), data = airquality) +
  geom_point(aes(x = Day, y = Wind), data = airquality)
```

```{r}
equiv_plot(p_data_global, p_data_local, check = c("layers", "aes"))
```

`spec_layers()$data_source` is `"global"` vs `"local"` in these two plots, but
that column is intentionally excluded from all `equiv_*` comparisons. Where data
lives is a stylistic choice that should not penalise a student who achieves the
same visual result.

The same holds for a two-dataset plot where the two layers draw from different
data frames (e.g. a base map layer and a data overlay). As long as the geom names
and aesthetic mappings of each layer match, `equiv_layers()` and `equiv_aes()`
pass regardless of how the data frames are distributed across `ggplot()` and each
`geom_*()` call.

### 1.3 Layer order

`equiv_layers(exact = FALSE)` (the default) performs a subset check on geom
names, so layer order does not matter. `compare_plots(mode = "structural")`
additionally sorts layers into a canonical order before comparison, making
`exact = TRUE` checks order-insensitive too.

```{r}
p_point_line  <- ggplot(airquality, aes(Day, Wind)) + geom_point() + geom_line()
p_line_point  <- ggplot(airquality, aes(Day, Wind)) + geom_line()  + geom_point()

# Subset check: passes regardless of order
equiv_layers(p_point_line, p_line_point)

# Exact check: fails without canonicalisation
equiv_layers(p_point_line, p_line_point, exact = TRUE)

# Structural mode sorts layers, so exact check passes
compare_plots(p_point_line, p_line_point,
              mode = "structural", check = "layers")
```

---

## 2. Bar chart equivalence

### 2.1 Direct equivalents

Global vs local `aes()` for `geom_bar()` is the same inheritance story as above:

```{r}
p_ref <- ggplot(mpg, aes(x = class)) + geom_bar()
p_loc <- ggplot(mpg) + geom_bar(aes(x = class))

equiv_plot(p_ref, p_loc, check = c("layers", "aes"))
```

Pre-counted data with `geom_bar(stat = "identity")` also passes `equiv_plot()`
directly. `equiv_layers(exact = FALSE)` only tests geom names ("bar" in both);
`equiv_aes(exact = FALSE)` uses subset matching, so the extra `y = n` mapping in
the observation does not fail the reference's `x = class` requirement:

```{r}
counts <- count(mpg, class)

p_identity <- ggplot(counts, aes(x = class, y = n)) +
  geom_bar(stat = "identity") + labs(y = "n")

equiv_plot(p_ref, p_identity, check = c("layers", "aes"))
```

### 2.2 `geom_col()` — structural mode required

`geom_col()` is internally `GeomCol`, whose name is `"col"`, not `"bar"`.
`equiv_layers()` therefore fails when the reference uses `geom_bar()`:

```{r}
p_col <- ggplot(counts, aes(x = class, y = n)) + geom_col() + labs(y = "n")

# Direct comparison fails on the layer check:
equiv_plot(p_ref, p_col, check = "layers")

# Structural mode applies .rule_geom_col_to_bar, normalising "col" → "bar":
compare_plots(p_ref, p_col, mode = "structural", check = "layers")
```

The same applies to any `geom_col()` combination regardless of whether
`scale_*` or `labs()` is also present.

### 2.3 `coord_flip()` — visual mode required

A bar chart written as `aes(y = class) + coord_flip()` has its x/y aesthetics
swapped relative to the canonical `aes(x = class)` form. `equiv_aes()` sees
`"1::x"` in the reference but only `"1::y"` in the observation, so it fails:

```{r}
p_flip <- ggplot(mpg, aes(y = class)) + geom_bar() + coord_flip()

# Direct aes check fails:
equiv_plot(p_ref, p_flip, check = "aes")

# Visual mode applies .rule_coord_flip, swapping x ↔ y and replacing
# coord_flip with coord_cartesian before comparison:
compare_plots(p_ref, p_flip, mode = "visual", check = c("layers", "aes", "coord"))
```

This applies equally to `geom_col() + coord_flip()` and to pre-counted variants.
All such plots require `mode = "visual"` when compared to a non-flipped reference.

### 2.4 Scale `name` vs `labs()` — visual mode required

`scale_y_continuous(name = "count")` stores the label inside the scale object;
`labs(y = "count")` stores it in `p$labels`. They appear different to
`equiv_labels()` unless visual mode is active:

```{r}
p_scale_name <- ggplot(counts, aes(x = class, y = n)) +
  geom_bar(stat = "identity") + scale_y_continuous(name = "count")

p_labs <- ggplot(counts, aes(x = class, y = n)) +
  geom_bar(stat = "identity") + labs(y = "count")

# Visual mode applies .rule_scale_name_to_labels, promoting the scale name
# into the labels table:
compare_plots(p_labs, p_scale_name, mode = "visual", check = "labels")
```

---

## 3. Count plot equivalence

### 3.1 `geom_count()` variants — direct equivalents

All coding styles for the same `geom_count()` plot are equivalent via standard
inheritance resolution:

```{r}
p_count_global <- ggplot(mpg, aes(x = drv, y = class)) + geom_count()
p_count_local  <- ggplot(mpg) + geom_count(aes(x = drv, y = class))
p_count_split  <- ggplot(mpg, aes(x = drv)) + geom_count(aes(y = class))

equiv_plot(p_count_global, p_count_local, check = c("layers", "aes"))
equiv_plot(p_count_global, p_count_split, check = c("layers", "aes"))
```

### 3.2 `geom_point(aes(size = n))` + pre-counted data — separate equivalence group

`geom_count()` uses `GeomCount` (geom name `"count"`). `geom_point()` uses
`GeomPoint` (geom name `"point"`). These are different names, so
`equiv_layers()` fails when comparing across the two approaches:

```{r}
mpg_counts <- count(mpg, drv, class)

p_point_sized <- ggplot(mpg_counts, aes(x = drv, y = class, size = n)) +
  geom_point()

# Fails: reference has geom "count", observation has geom "point"
equiv_plot(p_count_global, p_point_sized, check = "layers")
```

These two plots are visually equivalent but **not spec-equivalent** under the
current function set. They form separate equivalence groups:

- **Group A** — `geom_count()` variants: equivalent within the group via `equiv_plot()`.
- **Group B** — `geom_point(aes(size = n))` + `count()` variants: equivalent
  within the group via `equiv_plot()`.

Comparing across groups requires a custom canonicalisation rule that rewrites
both the geom and the data simultaneously. See **Section 5**.

---

## 4. Histogram binning variants

All histogram calls share geom name `"bar"` (histogram uses `GeomBar` +
`StatBin`), so `equiv_layers()` passes regardless of binning parameters.
`equiv_aes()` compares the `x` aesthetic, which is the same:

```{r}
p_hist_default  <- ggplot(mpg, aes(x = hwy)) + geom_histogram()
p_hist_bins30   <- ggplot(mpg, aes(x = hwy)) + geom_histogram(bins = 30)
p_hist_binwidth <- ggplot(mpg, aes(x = hwy)) + geom_histogram(binwidth = 3)

equiv_plot(p_hist_default, p_hist_bins30,   check = c("layers", "aes"))
equiv_plot(p_hist_default, p_hist_binwidth, check = c("layers", "aes"))
```

Binning parameters sit in the `params` list-column of `spec_layers()`. They can
be checked explicitly with `equiv_params()` when the grader wants to enforce a
specific binning choice:

```{r}
# Reference has no explicit bins; observation has bins = 30 — params differ
equiv_params(p_hist_default, p_hist_bins30, layer = 1L, params = "bins")
```

The `pedagogical` mode logs `bins` vs `binwidth` usage via
`.rule_histogram_bin_param`, without converting between them (the numeric values
are not inter-derivable without knowing the data range).

---

## 5. Pending work

### 5.1 Cross-geom stat equivalence

The `geom_count()` / `geom_point(aes(size = n))` + `count()` boundary is an
instance of a broader class: a plot that uses a stat-computing geom on raw data
vs a plot that pre-computes the same statistic and applies a simpler geom.
Current canonicalisation rules operate only on the spec (geom name, aes
mappings, coord, scale names). A rule that could bridge these groups would need
to:

1. Detect a stat-computing layer (`stat = "sum"` for `geom_count`, `stat = "bin"`
   for `geom_histogram`, etc.).
2. Recognise that the comparison plot provides pre-computed data whose columns
   match what the stat would have computed.
3. Rewrite the spec (and possibly verify the data) to produce a common form.

This requires integrating `equiv_data()` into the canonicalisation pipeline, which
is more involved than the existing rules. See `next.md`.

### 5.2 Multi-dataset layer verification

`spec_layers()$data_source` marks which layers carry their own data frame, and
`layer_data_index()` locates a specific data frame within a plot. For the
comparison to verify that *corresponding* layers in two plots draw from
semantically equivalent data frames, a matching step is needed. The current
`equiv_data()` function compares by hash (after sorting columns), but it
operates on a single layer index, not on matched pairs across plots. See `next.md`.
```
