---
title: "Plot Grading and Testing with ggspec"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plot Grading and Testing with ggspec}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Introduction

`ggspec` provides a comparison tier (`equiv_*()`) and a check/assertion tier
(`check_plot()`, `expect_equiv_plot()`) for comparing two ggplot objects. These
are designed to be framework-agnostic: they work in plain R scripts, `testthat`
test suites, and `learnr`/`gradethis` grading pipelines.

```{r}
library(ggspec)
library(ggplot2)
```

## Comparing two plots with `equiv_plot()`

`equiv_plot()` is the high-level entry point. It accepts two ggplot objects and
a character vector of check names to run. It returns a `ggspec_result` object
that holds a pass/fail flag, a human-readable message, and a structured diff.

```{r}
ref <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  facet_wrap(~drv) +
  labs(title = "Reference plot")

obs_correct <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  facet_wrap(~drv) +
  labs(title = "Reference plot")

obs_wrong <- ggplot(mpg, aes(displ, hwy)) +
  geom_smooth() +            # wrong geom
  facet_wrap(~cyl) +         # wrong facet variable
  labs(title = "Student plot")
```

```{r}
# Passing case
result_ok <- equiv_plot(ref, obs_correct)
result_ok
as.logical(result_ok)
```

```{r}
# Failing case
result_fail <- equiv_plot(ref, obs_wrong)
result_fail
```

### Running individual checks

Each `equiv_*()` function tests one dimension:

```{r}
equiv_layers(ref, obs_wrong)
equiv_facets(ref, obs_wrong)
equiv_labels(ref, obs_wrong, aesthetics = "title")
```

### The `exact` argument

By default, `equiv_layers()` and `equiv_aes()` use subset matching: the
observed plot must contain *at least* the layers/mappings of the reference.
Set `exact = TRUE` to require an exact match.

```{r}
obs_extra <- ref + geom_smooth()  # extra layer is fine by default
equiv_layers(ref, obs_extra)

equiv_layers(ref, obs_extra, exact = TRUE)  # fails: extra layer
```

## Framework-agnostic checking with `check_plot()`

`check_plot()` wraps `equiv_plot()` and calls a `fail_fn` if the check fails.
The default `fail_fn = stop` makes it work anywhere.

```{r error = TRUE}
# Passes silently
check_plot(obs_correct, ref, check = c("layers", "aes", "facets"))

# Fails with an informative error
check_plot(obs_wrong, ref, check = c("layers", "facets"))
```

### Swapping in `gradethis::fail()`

In a `learnr` tutorial, swap the `fail_fn`:

```r
# Inside a learnr grade_this() block:
check_plot(
  .result,
  expected = ref,
  check    = c("layers", "aes", "facets"),
  fail_fn  = gradethis::fail,
  pass_fn  = gradethis::pass
)
```

No hard dependency on `gradethis` is required.

### Using `expect_equiv_plot()` in `testthat`

```r
testthat::test_that("student plot has correct layers and facets", {
  expect_equiv_plot(
    obs_correct,
    ref,
    check = c("layers", "aes", "facets")
  )
})
```

## Inspecting the diff

Every `equiv_*()` result carries a `$detail` data frame for programmatic
inspection:

```{r}
result <- equiv_aes(ref, obs_wrong)
result$detail
```

## Comparing layer parameters

`equiv_params()` checks whether a specific layer's non-aesthetic parameters
match, e.g. checking that a student used `se = FALSE` on `geom_smooth()`.

```{r}
p_ref   <- ggplot(mpg, aes(displ, hwy)) + geom_smooth(method = "lm", se = FALSE)
p_wrong <- ggplot(mpg, aes(displ, hwy)) + geom_smooth(method = "lm", se = TRUE)

equiv_params(p_ref, p_wrong, layer = 1L, params = "se")
```

## Summary of available checks

| Function | What it checks |
|---|---|
| `equiv_layers()` | Geom and stat per layer |
| `equiv_aes()` | Aesthetic-to-variable mappings |
| `equiv_scales()` | Explicitly added scales |
| `equiv_facets()` | Facet type and variables |
| `equiv_labels()` | Title, axis, and aesthetic labels |
| `equiv_coord()` | Coordinate system type |
| `equiv_params()` | Non-aesthetic layer parameters |
| `equiv_data()` | Data hash per layer |
| `equiv_plot()` | All of the above in one call |
